---
title: "Introduction to SpatialDataR"
author: "Cuiwei Gao"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Introduction to SpatialDataR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Introduction

`SpatialDataR` provides native R infrastructure for reading
SpatialData-formatted Zarr stores without Python dependencies.
The Python `spatialdata` library
(Marconato et al. 2024 *Nature Methods*) defined a universal
Zarr-based format for spatial omics. This package bridges that
format to the R/Bioconductor ecosystem.

## Design principles

- **No Python required**: reads `.zarr` stores directly using
  R packages (`Rarr`, `pizzarr`, `arrow`)
- **SpatialData-aware**: parses `.zattrs` metadata, discovers
  elements, handles coordinate systems
- **Bioconductor-native**: returns `DataFrame`, `SimpleList`,
  and `SpatialExperiment` objects
- **Format-flexible**: supports CSV and Parquet for tabular
  elements, Zarr arrays for images/labels

## Comparison with alternatives

| Feature | SpatialDataR | reticulate | Rarr/pizzarr |
|---------|:---:|:---:|:---:|
| Python-free | ✓ | ✗ | ✓ |
| SpatialData metadata | ✓ | ✓ | ✗ |
| Coordinate transforms | ✓ | ✓ | ✗ |
| Bioconductor objects | ✓ | ✗ | ✗ |

# Installation

```{r eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("SpatialDataR")
```

# Quick start

## Reading a SpatialData store

The package ships a minimal Zarr store (`xenium_mini.zarr`)
following the SpatialData specification. It contains a
simulated 10x Xenium dataset: 50 cells, 10 genes, 500
transcripts.

```{r read-store}
library(SpatialDataR)
library(S4Vectors)

store_path <- system.file("extdata", "xenium_mini.zarr",
    package = "SpatialDataR")
sd <- readSpatialData(store_path)
sd
```

## Accessing elements

Each element type has a dedicated accessor. Images and labels
are stored as path references (use `readZarrArray()` to load
the actual data). Points, shapes, and tables are loaded
eagerly from CSV/Parquet.

```{r accessors}
# Image references (path + metadata)
images(sd)

# Labels references
spatialLabels(sd)

# Points — loaded as DataFrame
pts <- spatialPoints(sd)[["transcripts"]]
head(pts)

# Shapes — loaded as DataFrame
shp <- shapes(sd)[["cell_boundaries"]]
head(shp)

# Tables — obs/var metadata
tbl <- tables(sd)[["table"]]
head(tbl$obs)
```

## Selective reading

Read only specific element types for faster loading:

```{r selective}
sd_img <- readSpatialData(store_path,
    elements = c("images", "labels"))
sd_img
```

# Element-level readers

## CSV elements

Many SpatialData stores include CSV-backed tabular data:

```{r csv-reader}
csv_path <- file.path(store_path, "points", "transcripts",
    "transcripts.csv")
pts_df <- readCSVElement(csv_path)
head(pts_df)
```

## Tables (AnnData format)

`readSpatialTable()` reads the AnnData-formatted table group,
extracting obs (cell metadata) and var (gene metadata):

```{r table-reader}
tbl <- readSpatialTable(
    file.path(store_path, "tables", "table"))
str(tbl)
```

## Zarr arrays

For image and label data stored as Zarr arrays, use
`readZarrArray()` (requires `Rarr` or `pizzarr`):

```{r zarr-reader, eval=FALSE}
img <- readZarrArray(
    file.path(store_path, "images", "morphology", "scale0"))
dim(img)
```

# Coordinate transformations

SpatialData stores define coordinate systems for aligning
multi-modal data. `SpatialDataR` supports affine transforms
for mapping between coordinate spaces.

## Creating transforms

```{r transforms}
# Identity (no-op)
ct_id <- CoordinateTransform("identity")

# Pixel-to-micron transformation (0.2125 µm/pixel)
ct <- CoordinateTransform("affine",
    affine = matrix(c(0.2125, 0, 0,
                      0, 0.2125, 0,
                      0, 0, 1), nrow = 3, byrow = TRUE),
    input_cs = "pixels", output_cs = "microns")
ct
```

## Applying transforms

Works on both `DataFrame` and `matrix` inputs:

```{r apply-transform}
# DataFrame
pts <- DataFrame(x = c(100, 200, 300), y = c(50, 150, 250))
pts_um <- transformCoords(pts, ct)
pts_um

# Matrix
coords <- matrix(c(100, 200, 50, 150), ncol = 2)
colnames(coords) <- c("x", "y")
coords_um <- transformCoords(coords, ct)
coords_um
```

## Parsing transforms from metadata

Element metadata often contains coordinate transformation
definitions. The internal `.parseTransform()` function handles
identity, affine, and scale types:
```{r parse-transform}
img_meta <- images(sd)[["morphology"]]$metadata
img_meta$coordinateTransformations
```

# Coordinate systems

The top-level `.zattrs` defines the coordinate systems for
aligning multi-modal elements:

```{r coord-systems}
cs <- coordinateSystems(sd)
str(cs)
```

# Session info

```{r session}
sessionInfo()
```

# References

Marconato L et al. (2024). SpatialData: an open and universal
data framework for spatial omics. *Nat Methods* 21:2196-2209.
doi:10.1038/s41592-024-02212-x
