---
title: "Introduction to SpatialDataR"
author: "Cuiwei Gao"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Introduction to SpatialDataR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Introduction

`SpatialDataR` provides native R infrastructure for reading
SpatialData-formatted Zarr stores (Marconato et al. 2025
*Nature Methods*) without Python dependencies. The package
implements the data model described in the SpatialData
specification, which builds on the OME-NGFF standard (Moore
et al. 2023 *Histochem Cell Biol*).

## Design principles

- **No Python required**: reads `.zarr` stores using
    R packages (`Rarr`, `pizzarr`, `arrow`, `jsonlite`)
- **SpatialData-aware**: parses `.zattrs` metadata,
    discovers elements, handles coordinate systems
- **Bioconductor-native**: returns `DataFrame`,
    `SimpleList`, and `SpatialExperiment` objects
- **OME-NGFF compliant**: supports identity, scale,
    translation, affine, and sequence transforms

## Comparison with alternatives

| Feature | SpatialDataR | reticulate | Rarr/pizzarr |
|---------|:---:|:---:|:---:|
| Python-free | yes | no | yes |
| SpatialData metadata | yes | yes | no |
| Coordinate transforms | yes | yes | no |
| Spatial queries | yes | yes | no |
| Bioconductor objects | yes | no | no |

# Quick start

```{r read-store}
library(SpatialDataR)
library(S4Vectors)

store_path <- system.file("extdata", "xenium_mini.zarr",
    package = "SpatialDataR")
sd <- readSpatialData(store_path)
sd
```

## Element summary

```{r element-summary}
elementSummary(sd)
```

## Accessing elements

Images and labels are stored as path references (use
`readZarrArray()` to load). Points and shapes are loaded
eagerly from CSV/Parquet as `DataFrame` objects.

```{r accessors}
# Points (loaded as DataFrame)
pts <- spatialPoints(sd)[["transcripts"]]
head(pts)

# Shapes
shp <- shapes(sd)[["cell_boundaries"]]
head(shp)

# Tables (obs/var metadata)
tbl <- tables(sd)[["table"]]
head(tbl$obs)

# Image reference
images(sd)[["morphology"]][c("name", "type")]
```

## Selective reading

```{r selective}
sd_pts <- readSpatialData(store_path,
    elements = c("points", "shapes"))
elementSummary(sd_pts)
```

## Subsetting by element name

```{r subset}
sub <- sd["transcripts"]
sub
```

# Coordinate transformations

SpatialData stores define coordinate systems for aligning
multi-modal data. `SpatialDataR` supports the full OME-NGFF
transformation vocabulary.

## Creating transforms

```{r transforms-basic}
# Pixel-to-micron scaling (Xenium: 0.2125 um/pixel)
ct <- CoordinateTransform("affine",
    affine = diag(c(0.2125, 0.2125, 1)),
    input_cs = "pixels", output_cs = "microns")
ct
```

## Extracting transforms from elements

```{r element-transform}
# Get the transform from image metadata
ct_img <- elementTransform(images(sd)[["morphology"]])
ct_img
```

## Applying transforms

```{r apply-transform}
# DataFrame
pts_sample <- DataFrame(x = c(100, 200, 300),
    y = c(50, 150, 250))
pts_um <- transformCoords(pts_sample, ct)
pts_um

# Matrix (also supports 3D: Nx3)
coords <- matrix(c(100, 200, 50, 150), ncol = 2)
colnames(coords) <- c("x", "y")
transformCoords(coords, ct)
```

## Composing transforms

Chain a scale and a translation into a single transform:

```{r compose}
scale_ct <- CoordinateTransform("affine",
    affine = diag(c(0.5, 0.5, 1)),
    input_cs = "raw", output_cs = "scaled")

translate_ct <- CoordinateTransform("affine",
    affine = matrix(c(1,0,100, 0,1,200, 0,0,1),
        nrow = 3, byrow = TRUE),
    input_cs = "scaled", output_cs = "global")

combined <- composeTransforms(scale_ct, translate_ct)
combined

# Equivalent to: first scale by 0.5, then shift by (100, 200)
pts_test <- DataFrame(x = c(10), y = c(20))
transformCoords(pts_test, combined)
```

## Inverting transforms

Map back from physical to pixel coordinates:

```{r invert}
inv <- invertTransform(ct)
inv

# Roundtrip: pixels -> microns -> pixels
pts_orig <- DataFrame(x = c(100), y = c(200))
pts_fwd <- transformCoords(pts_orig, ct)
pts_back <- transformCoords(pts_fwd, inv)
pts_back  # Should match pts_orig
```

# Spatial queries

`bboxQuery()` subsets elements to a rectangular region,
mirroring Python's `spatialdata.bounding_box_query()`.

```{r bbox-query}
# Full dataset
nrow(spatialPoints(sd)[["transcripts"]])

# Subset to [0, 2] x [0, 2] microns
sub_sd <- bboxQuery(sd, xmin = 0, xmax = 2,
    ymin = 0, ymax = 2)
nrow(spatialPoints(sub_sd)[["transcripts"]])
```

```{r bbox-dataframe}
# Also works directly on DataFrames
sub_pts <- bboxQuery(pts, xmin = 0, xmax = 1,
    ymin = 0, ymax = 1)
nrow(sub_pts)
```

# Coordinate systems

```{r coord-systems}
cs <- coordinateSystems(sd)
str(cs)
```

# Session info

```{r session}
sessionInfo()
```

# References

- Marconato L et al. (2025). SpatialData: an open
    and universal data framework for spatial omics.
    *Nat Methods* 22:58-62.
    doi:10.1038/s41592-024-02212-x

- Moore J et al. (2023). OME-Zarr: a cloud-optimized
    bioimaging file format with international community
    support. *Histochem Cell Biol* 160:223-251.
    doi:10.1007/s00418-023-02209-1

- Righelli D et al. (2022). SpatialExperiment:
    infrastructure for spatially-resolved
    transcriptomics data in R using Bioconductor.
    *Bioinformatics* 38:3128-3131.
    doi:10.1093/bioinformatics/btac299
