---
title: "Introduction to SpatialDataR"
author: "CuiweiG"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Introduction to SpatialDataR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Introduction

`SpatialDataR` provides native R access to SpatialData Zarr stores
without Python dependencies. The Python `spatialdata` library
(Marconato et al. 2024 *Nature Methods*) defined a universal format
for spatial omics — but R users had no way to read it without
`reticulate`. This package fills that gap.

## Why not reticulate?

Using Python bridges introduces:

- Conda/virtualenv dependency management
- Type conversion overhead
- Loss of lazy loading (DelayedArray)
- No native SpatialExperiment integration

`SpatialDataR` reads Zarr directly using `Rarr`/`pizzarr` and
returns Bioconductor-native objects.

# Quick start

## Reading a SpatialData store

```{r mock-store}
library(SpatialDataR)

# Create a mock SpatialData Zarr store for demonstration
tmp <- tempfile()
dir.create(file.path(tmp, "images", "morphology"), recursive = TRUE)
dir.create(file.path(tmp, "points", "transcripts"), recursive = TRUE)
dir.create(file.path(tmp, "tables", "adata"), recursive = TRUE)
dir.create(file.path(tmp, "labels", "cells"), recursive = TRUE)

# Write metadata (SpatialData spec)
writeLines('{"spatialdata_attrs": {"version": "0.1"}}',
    file.path(tmp, ".zattrs"))
writeLines('{}', file.path(tmp, "images", "morphology", ".zattrs"))
writeLines('{}', file.path(tmp, "points", "transcripts", ".zattrs"))
writeLines('{}', file.path(tmp, "tables", "adata", ".zattrs"))
writeLines('{}', file.path(tmp, "labels", "cells", ".zattrs"))

# Read the store
sd <- readSpatialData(tmp)
sd
```

## Accessing elements

```{r accessors}
# Each element type has an accessor
images(sd)
spatialLabels(sd)
spatialPoints(sd)
shapes(sd)
tables(sd)
coordinateSystems(sd)
```

## Selective reading

```{r selective}
# Read only specific element types
sd_img <- readSpatialData(tmp, elements = c("images", "labels"))
sd_img
```

```{r cleanup-store, include=FALSE}
unlink(tmp, recursive = TRUE)
```

# Coordinate transformations

SpatialData stores include coordinate system definitions for
aligning multi-modal data. `SpatialDataR` supports affine
transformations natively.

```{r transforms}
library(S4Vectors)

# Create point coordinates (in pixel space)
pts <- DataFrame(x = c(100, 200, 300, 400),
                 y = c(50, 150, 250, 350))

# Define pixel-to-micron transformation (0.5 um/pixel + offset)
pixel_to_um <- CoordinateTransform("affine",
    affine = matrix(c(0.2125, 0, 10,
                      0, 0.2125, 20,
                      0, 0, 1), nrow = 3, byrow = TRUE),
    input_cs = "pixels",
    output_cs = "microns")
pixel_to_um

# Apply transformation
pts_um <- transformCoords(pts, pixel_to_um)
pts_um
```

# Architecture

SpatialDataR maps SpatialData elements to Bioconductor objects:

| SpatialData element | Zarr format | R representation |
|---------------------|-------------|------------------|
| images | Zarr arrays | DelayedArray (lazy) |
| labels | Zarr arrays | DelayedArray (lazy) |
| points | Parquet | DataFrame |
| shapes | GeoParquet | DataFrame |
| tables | AnnData/Zarr | SpatialExperiment |

All reading is lazy by default — data stays on disk until
explicitly accessed.

# Working with real data

To analyze a real Xenium or Visium dataset stored in SpatialData
format:

```r
# Read a 10x Xenium dataset
sd <- readSpatialData("xenium_breast_cancer.zarr")

# Get the expression table as SpatialExperiment
spe <- readSpatialTable("xenium_breast_cancer.zarr/tables/table")

# Read transcript coordinates
pts <- readParquetPoints("xenium_breast_cancer.zarr/points/transcripts")

# Transform to physical coordinates
ct <- coordinateSystems(sd)[["global"]]
pts_um <- transformCoords(pts, ct)
```

# Session info

```{r session}
sessionInfo()
```
