---
title: "Introduction to SpatialDataR"
author: "CuiweiG"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Introduction to SpatialDataR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Installation

```{r eval=FALSE}
if (!requireNamespace("remotes", quietly = TRUE))
    install.packages("remotes")
remotes::install_github("CuiweiG/SpatialDataR")
```

# Introduction

`SpatialDataR` provides native R infrastructure for reading
SpatialData-formatted Zarr stores without Python dependencies.
The Python `spatialdata` library (Marconato et al. 2024 *Nature
Methods*) defined a universal format for spatial omics.
R-side support has remained limited; this package provides a
lightweight bridge to the Bioconductor ecosystem.

## Positioning

`SpatialDataR` complements existing tools:

- **vs reticulate**: zero Python dependency, native R objects
- **vs Rarr/pizzarr**: SpatialData-aware metadata parsing,
  not just raw Zarr array reading
- **vs SpatialExperiment**: reads SpatialData *stores*, not
  individual matrices

# Reading a SpatialData store

The package ships a minimal Zarr store (`xenium_mini.zarr`)
following the SpatialData specification.

```{r read-store}
library(SpatialDataR)

store_path <- system.file("extdata", "xenium_mini.zarr",
    package = "SpatialDataR")
sd <- readSpatialData(store_path)
sd
```

## Accessors

```{r accessors}
images(sd)
spatialLabels(sd)
spatialPoints(sd)
shapes(sd)
tables(sd)
```

## Coordinate systems

```{r coord-sys}
coordinateSystems(sd)
```

## Selective reading

```{r selective}
sd_img <- readSpatialData(store_path,
    elements = c("images", "labels"))
sd_img
```

# Coordinate transformations

SpatialData stores define coordinate systems for aligning
multi-modal data. `SpatialDataR` supports affine transforms.

```{r transforms}
library(S4Vectors)

pts <- DataFrame(x = c(100, 200, 300), y = c(50, 150, 250))

# Pixel-to-micron transformation (0.2125 um/pixel)
ct <- CoordinateTransform("affine",
    affine = matrix(c(0.2125, 0, 0,
                      0, 0.2125, 0,
                      0, 0, 1), nrow = 3, byrow = TRUE),
    input_cs = "pixels", output_cs = "microns")
ct

pts_um <- transformCoords(pts, ct)
pts_um
```

# Session info

```{r session}
sessionInfo()
```
